---
title: "Reproducible Research - Peer Assesment 1"
output: html_document
---

**Load and preprocessing the Data**

The Activity Monitoring data for this analysis can be downloaded using this link (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

Save this file to your working directory then unzip and open the file. 

Load lattice package

```{r}
unzip('repdata-data-activity.zip')
activitydata <- read.csv("activity.csv")
head(activitydata)

library(lattice)
```

Use as.date to convert character data to dates

```{r}
activitydata$date <- as.Date(activitydata$date)
```

**What is mean total number of steps taken per day?**

Find the total number of steps ignoring any missing values and create a Histogram of total number of steps taken each day

```{r}
StepsTotal <- aggregate(steps ~ date, data = activitydata, sum, na.rm=TRUE)
hist(StepsTotal$steps, main = "Total Steps By Day", xlab = "Days", col = 'green')
```

Calculate and report the Mean and Median of the total number of steps taken each day

```{r}
mean(StepsTotal$steps, trim = 0.5)
median(StepsTotal$steps)
```

**What is the average daily activity pattern?**

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
timeseries <- tapply(activitydata$steps, activitydata$interval, mean, na.rm=TRUE)
plot(row.names(timeseries), timeseries, type = "l", xlab = "5-min interval", ylab = "Average across all days", main = "Average number of steps taken", col='green')
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxinterval <- which.max(timeseries)
names(maxinterval)
```

**Imputing missing values**

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
activitydataNA <- sum(is.na(activitydata$steps))
activitydataNA
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc

I will use the means for the 5-minute intervals as fillers for missing values.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
StepsAverage <- aggregate(steps ~ interval, data = activitydata, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(activitydata)) {
    obs <- activitydata[i, ]
    if (is.na(obs$steps)) {
        steps <- subset(StepsAverage, interval == obs$interval)$steps
    } else {
        steps <- obs$steps
    }
    fillNA <- c(fillNA, steps)
}
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
new_activity <- activitydata
new_activity$steps <- fillNA
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
StepsTotalupdated <- aggregate(steps ~ date, data = new_activity, sum, na.rm=TRUE)
hist(StepsTotalupdated$steps, main = "Total Steps By Day", xlab = "Days", col = 'green')
mean(StepsTotalupdated$steps)
median(StepsTotalupdated$steps)
```

The orignal and updated median is shown in the table below. The difference in each case is only 1 step.  

***Original***

Mean 10765 

Median 10765

***Updated***

Mean 10766

Median 10766

**Are there differences in activity patterns between weekdays and weekends?**

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
day <- weekdays(new_activity$date)
daylevel <- vector()
        for (i in 1:nrow(new_activity)){
            if (day[i] == "Saturday"){
                daylevel[i] <- "Weekend"
            } else if (day[i] == "Sunday"){
                daylevel[i] <- "Weekend"
            } else {
                daylevel[i] <- "Weekday"
            }
        }
new_activity$daylevel <- daylevel
new_activity$daylevel <- factor(new_activity$daylevel)

stepsByDay <- aggregate(steps ~ interval + daylevel, data = new_activity, mean)
names(stepsByDay) <- c("interval", "daylevel", "steps")
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
xyplot(steps ~ interval | daylevel, stepsByDay, type = "l", layout = c(1, 2), xlab = "Interval", ylab = "Number of steps")
```

